---
title: "Weather data exploration"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

```{r}
library(tidyverse)
library(janitor)
library(lubridate)
library(here)
```

```{r}
source(here("data_cleaning_scripts/flight_delays.R"))
```
# Data exploration 
```{r}
weather %>% 
  summarise(
    across(
      everything(),
      ~sum(is.na(.))
    ))
```
Temp, dewp, humid, precip, pressure all have significant number of missing values. Will search for other datasets with this info. Dropping from this dataset and beginning cleaning process/  

```{r}
test_weather_clean <- weather %>%
  select(-c(year, month, day, hour, temp,
            dewp, humid, precip, pressure))
```

```{r}
test_weather_clean %>% 
  summarise(
    across(
      everything(),
      ~sum(is.na(.))
    ))
```
Still over 200 missing values in wind_dir, wind_speed and wind_gust. Will merge to flights data to see how many flights this results in with missing weather data before decided what to do. 

Visibility has only 11 missing values. Will drop these 11 rows.

```{r}
# add to cleaning script
test_weather_clean %>% 
  drop_na(visib)
```

Reviewing NAs in wind_dir

```{r}
test_weather_clean %>% 
  filter(is.na(wind_dir)) %>% 
  distinct(visib)
```

```{r}
weather_clean %>% 
  filter(visib == 10) %>% 
  distinct(wind_speed)
```
Opting to drop NAs from wind_dir, wind_speed and wind_gust

```{r}
# add to cleaning script
weather_clean %>% 
  drop_na(wind_dir) %>% 
  drop_na(wind_speed) %>% 
  drop_na(wind_gust)
```

```{r}
weather_clean %>% 
  summarise(
    across(
      everything(),
      ~sum(is.na(.))
    ))
```
All NAs dealt with. 

```{r}
# checking data contains info for all 3 airports
weather_clean %>% 
  distinct(origin)
```

```{r}
#checking data contains all of 2017
time_hour <- weather_clean %>% 
  distinct(time_hour)
# confirmed
```

```{r}
test_join <- flights_clean %>% 
              left_join(weather_clean, 
                        by = c("origin", "time_hour"))
```

```{r}
test_join %>% 
  filter(origin == "EWR",
         time_hour == "2017-10-01 21:00:00")
```
```{r}
weather_clean %>% 
  filter(origin == "EWR",
         time_hour == "2017-10-01 21:00:00")
```
Joining tested, spotted checked and functioning correctly.

```{r}
#checking NAs in newly joined data

flight_weather %>% 
  summarise(
    across(
      everything(),
      ~sum(is.na(.))
    ))
```
24686 missing values from wind_dir, wind_speed, wind_gust, visib.

```{r}
# how much of data is missing?
24686/303748 *100

#how much data will remain if drop NAs
303748 - 24686
```
If we drop missing values we will lose 8.1% of the data. However, the alternative is inserting a statistical calculation for wind information which could lead to inaccuracies in the final model.

As our business brief states that the company believe poor weather conditions are causing too many delays and want to invest in improving facilities, so that aircraft can take off in more types of weather, we do not want to assume data related to weather. Dropping the missing values will leave 27,9062 rows.

# Additional dataset

Sourced from: https://meteostat.net/en/station/72502?t=2017-01-01/2017-12-31

Information on avg_temp, precipitation and snow for every day in 2017. Will join to current dataset. Checking for NAs here before adding to cleaning script:

```{r}
additional_weather %>% 
  summarise(
    across(
      everything(),
      ~sum(is.na(.))
    )
  )
```
No NAs, adding and joining to weather_clean in cleaning script.


